Exploiting First-Class Arrays in Fortran for Accelerator Programming

### 2010-12-11 opencl/papers/foropencl.tex
Cleaned up after massive copy and paste operation.

%
% \section{Introduction}
%    \subsection{Approach}
%    \subsection{Why Fortran?}
%    \subsection{Other Languages}
%

%
% \section{Programming model}
%    \subsection{Fortran Syntax}   
%       \subsubsection*{Array notation}
%       \subsubsection*{Elemental functions}
%       \subsubsection*{Pure procedures}
%       \subsubsection*{Shift operators}
%       \subsubsection*{Regions}
%    \subsection{Parallelism}
%    \subsection{New Functions}
%    \subsection{Limitations}
%

@@

%% NEW - CER
However, the performance that these accelerator architectures offer comes at a
cost as processor architectures are trending toward multiple cores with
instances of integrated accelerator units and less of a reliance on memory
caches.  These changes place a heavy burden on application programmers as they
are forced to adapt to the new systems.  An especially challenging problem
faced by application programmers is not only how to program to these new
architectures (considering the massive scale of concurrency available), but
also how to design programs that are portable across the changing landscape of
computer architectures.  The fundamental question we address is what
programming model and what languages are best suited for these platforms.

@@

%% NEW - CER
The goal of this work is to exploit source-to-source transformations that
allow programmers to develop and maintain programs at a high-level of
abstraction, without coding to a specific hardware architecture.
Furthermore these transformations allow multiple hardware architectures
to be targeted without changing the high-level source.  It also removes the
necessity for application programmers to understand details of the accelerator
architecture or to know OpenCL.

We use the ROSE compiler infrastructure \cite{} to develop these
transformations.  ROSE uses the Open Fortran Parser \cite{} to parse Fortran
2008 syntax and can output (unparse) to OpenCL.  Since ROSE's intermediate
representation (IR) was constructed to represent multiple languages, it is
relatively straightforward to transform high-level Fortran IR nodes to OpenCL
nodes.  Furthermore, the Fortran array syntax maps directly to one, two, and
three-dimensional thread groups in OpenCL.

@@

We study automatic transformations of Fortran for an application example that
is typical of many applications that are based on finite-difference or
finite-volume methods in computational fluid dynamics (CFD).  The example is a
simple shallow water model in two dimensions using finite volume methods with
.... for time update.

%% NEW - CER
PAGOSA is a production code at LANL completely written in data-parallel
Fortran.  In a prior work we investigated automatically transforming this code
to run on LANL's Petaflop Roadrunner computer (a hybrid mixture of Opterons
and IBM Cell processors).  We demonstrated that a source-to-source compiler
can automatically vectorize and parallelize (both within the Cell processor
and across nodes) a small section of this code.  Preliminary results showed a
12 times performance gain on the Cell processor over a traditional single core
processor.  A factor of 9 is gained from vectorizing and parallelizing the
code to run on the Cell and the rest is gained from automatically overlapping
inter-node communication with computation on the Cell.

@@

%% NEW - CER
However, with the introduction of Fortran 90 (and later revisions to the
standard), Fortran is a truly modern programming language. Much of the arcane
baggage associated with it is actually no longer part of the language used by
modern programmers, and exists in a deprecated form simply to support legacy
codes. It is now modular and has many object-oriented features. Furthermore,
Fortran should be of interest to those studying parallel programming because
of its functional and data parallel constructs and because of the coarray
notation introduced after Fortran 2003. Unlike languages like C and C++, which
are considered to be more modern, Fortran has become a truly parallel language
with features added to recent language standards. With coarrays, Fortran
provides a mechanism to run replicas (called images) of a program on many
different processors (or cores) at the same time. It also provides language
constructs that allow the program to both read from and write to data residing
on {\it remote} processors and to synchronize the different program images so
that access to the data by the multiple images can be done atomically.

A likely question that one may pose is ``\emph{Why Fortran and not a more
  modern language like X?}''  The recent rise in interest in concurrency and
parallelism at the language level driven by multicore CPUs and manycore
accelerators has driven a number of new language developments, both as novel
languages and extensions on existing ones.

@@

%%
%%
%%

\section{The programming model and Fortran-language subset}

The static analysis and source-to-source transformations used are very
basic and simply require the programmer to use a subset of Fortran
that employs a data-parallel programming model.  In particular, it
encourages use of language features that were introduced and
standardized in the Fortran 90 language specification.  In this
section we describe the set of Fortran 90 features that our analysis
and transformation method are based on.

@@

%% NEW - CER
We will examine how the existing data-parallel constructs in Fortran can be
combined with coarrays or MPI to provide effectively a new parallel programming
language, one that is evolutionary in nature and provides complete compatibility
with existing applications and libraries.  Data parallelism is a high-level
abstraction that is, at the same time, both easier to program and gives the
compiler more leeway (if fully exploited) in retargeting a program to different
computer architectures.

@@

%% NEW - CER
The power in this notation is that the compiler is free to distribute the
computation in these expressions over any hardware processing elements available
to it, such as the vector units on an Intel or AMD processor or the Synergistic
Processing Units on an IBM Cell processor.  If the arrays are declared as
coarrays, this includes spreading the computation over \emph{all of the nodes in
  a cluster as well.}  In this case, communication occurs within the
\texttt{cshift} functions, though the compiler is free to overlap communication
and computation by scheduling the communication early and performing the
computation on the interior of the arrays while waiting for the communication to
complete.

%% NEW - CER
Complete and very concise and elegant programs can be built with statements
similar to that shown in Listing~\ref{lst:dpexample}, using intrinsic
functions like the array constructors (\texttt{CSHIFT}, \texttt{MERGE},
\texttt{TRANSPOSE}, ...), the array location functions (\texttt{MAXLOC} and
\texttt{MINLOC}), and the array reduction functions (\texttt{ANY},
\texttt{COUNT}, \texttt{MINVAL}, \texttt{SUM}, \texttt{PRODUCT},
...)~\cite{guydp}.  It should be emphasized, however, that {\it no} compiler
currently supplies coarray versions of these intrinsic functions as they are
not defined by the proposed Fortran 2008 standard (we have submitted a paper
proposing the integration of coarrays with these intricics functions to the J3
Fortran committee~\cite{rasmussen-caf-intrinsics}).  With coarray or MPI
implementations of these intrinsic functions, programs written in this
data-parallel subset of Fortran are implicitly parallel.  Furthermore, in
instances it is necessary to implement algorithms that don't lend themselves
to expressibility in this data-parallel subset, coarray notation can be used
for complete access to individual (including remote) data elements.  It is the
combination of data parallel and coarray constructs, with the transformational
properties of a source-to-source compiler (transforming array notation to
optimized Fortran 95 and library calls), that make this possibility such a
promising avenue for research.

%% NEW - CER
This style of programming meets the requirements we have set for a programming
model for developing petascale applications.  It allows the programmer to
program at a very-high level of abstraction while providing the compiler with
maximum flexibility in targeting the application for a particular hardware
architecture.  The data parallel programming model simultaneously meets the
seemingly conflicting goals of maintainability, portability, and performance.

@@

%% NEW - CER
The goal of this research is to demonstrate that parallel programs written in
this style of Fortran are maintainable and can achieve good performance on a
variety of processor architectures, including possibly, future terascale
architectures from Intel (such as the Larrabee architecture).  Using a code
segment from a production code at Los Alamos (written entirely in this style
of Fortran), we have demonstrated the potential for source-to-source
transformations to target the IBM Cell architecture used in Los Alamos
National Laboratories' Roadrunner computer~\footnote{Roadrunner is the first
  computer to break the petaflop barrier.}.  Initial results show that
transformed code using OFP and ROSE can achieve a speedup of up to 13 times on
sixteen Roadrunner nodes, compared with the equivalent number of single-core
Opteron nodes.

%% NEW - CER
The use of array notation allows one to program in a data-parallel subset of
Fortran. This style of programming makes use of array notation and pure and
elemental functions to operate on array elements.

%% NEW - CER
Because elemental functions return scalar values and are free from side
effects, the compiler is free to distribute the computation over any hardware
processing elements available to it, such as the multiple cores and vector
units on an Intel or AMD processor or the Synergistic Processing Units on an
IBM Cell processor.

%% NEW - CER
An example of an elemental function is shown in Listing~\ref{lst:elemental}.
This example simply returns its input divided by 2. While this may seem like a
trivial example, such simple functions may be composed with other elemental
functions to perform powerful computations, especially when applied to arrays.

Pure procedures, like elemental functions, must be free of side effects.
The absence of side effects removes ordering constraints from the compiler
allowing it to invoke pure functions out of order.  Procedures and functions
of this sort are also common in pure functional languages like Haskell,
and are exploited by compilers in order to emit parallel code automatically
due to their suitability for compiler-level analysis.

%% NEW - CER
The use of an elemental function is shown in the pure procedure {\tt laplace}
in Listing~\ref{lst:pure}. This function (for example) can be used to solve
the steady-state temperature along a one-dimensional surface. The temperature
at any point along the surface is equal to the average of the temperature at
neighboring points. Rather than writing explicit loops and using indexing to
reference neighboring points, one may simply shift the array {\tt T} to the
left and right, add the results and divide by 2. The intrinsic function {\tt
EOSHIFT} (end-off shift) semantically returns an array copy with the value at
any element shifted by the amount given by the {\tt shift} argument. The value
shifted in at the boundary is supplied by the {\tt boundary} argument. The
elemental function {\tt one\_half} is used to divide by 2.

%% NEW - CER
Note that semantically the elemental functions return their array results by
value. This is important because had loops been used to express the {\tt
laplace} procedure, temporary variables would have been required. Otherwise,
{\tt T(i)} would be updated with new values of {\tt T(i-1)}.

%% NEW - CER
 A ``pure'' function is one
without side effects (though it may change the value of array elements in an
array argument). An elemental function is a pure function with the additional
constraint that it takes only scalar arguments and must return a scalar.
Though defined in terms of scalars, surprisingly, an elemental function may
take an array as an actual argument, with the compiler applying the function
to each element of the array, and returning an array result. The plus operator
as described in the previous paragraph can be seen as an elemental function,
as the plus operator is defined in terms of scalars but can be applied to
whole arrays and can return an array result.

\subsubsection*{Shift operators}

%% NEW - CER
Arrays are first class citizens in Fortran as they are part of the language
type system rather than just a pointer to raw memory as in C. A Fortran array
contains metadata associated with it that describes the array's shape and
size. This metadata is lacking in other languages, such as C, C++, and Java.
Even in the case where multidimensional array data structures are available
(such as in C++ or Java), they exist outside the language standard, limiting
the ability of the compiler to analyze them fully. The existence of this
metadata allows the use of array notation whereby explicit indexing of arrays
is not required. For example, if {\tt A}, {\tt B}, and {\tt C} are all arrays
of the same rank and shape, then the statement {\tt C = A + B} results in the
sum of elements of {\tt A} and {\tt B} being stored in the corresponding
elements of {\tt C}. The first element of {\tt C} will contain the value of
the first element of {\tt A} added to the first element of {\tt B}.

The final piece that we exploit is the array notation present in modern
versions of Fortran.  For example,

\begin{equation}
\label{eq:dataparallel}
A = B + s * C
\end{equation}

is a array assignment statement where A, B, and C are arrays and s is a
scalar.  Note that no explicit iteration over array indices is needed and that
the individual operators, plus, times, and assignment are applied by the
compiler to individual elements of the arrays independently.  Thus the
compiler is able to spread the computation in~\ref{eq:dataparallel} across any
hardware threads under its control.

There are several advantages to this style of programming:

\begin{itemize}
	\item There are no loops or index variables to keep track of.  Off by one
index errors are a common programming mistake.
        \item The written code is closer to the algorithm, easier to understand, and is usually substantially shorter.
	\item Semantically the intrinsic functions return arrays by value.  This is usually what the algorithm requires.
	\item Because pure and elemental function are free from side effects, it is easier for a compiler to schedule the work to be done in parallel.
\end{itemize}

Complete and very concise and elegant programs can be built with procedures
similar to the {\tt laplace} example shown above. To aid this effort, Fortran
supplies intrinsic functions like the array constructors (\texttt{CSHIFT},
\texttt{EOSHIFT}, \texttt{MERGE}, \texttt{TRANSPOSE}, ...), the array location
functions (\texttt{MAXLOC} and \texttt{MINLOC}), and the array reduction
functions (\texttt{ANY}, \texttt{COUNT}, \texttt{MINVAL}, \texttt{SUM},
\texttt{PRODUCT}, ...).

\subsection{Relationship to other languages}

%% cites: ZPL, CM5 stencil compilers

% put something here on:
%
% shifts
% array notation
%

\subsubsection{elemental functions}

\subsubsection{pure procedures}

Mention that don't require pure procedures as region functions are not pure.
But programmers should otherwise think in terms of writing pure procedures.

@@

@@

This sections show short Fortran code examples and OpenCL equivalent.
The notation uses uppercase for arrays and lowercase for scalar quantities.
Interior array subsections are denoted by an i preceeding the array, e.g. iA
is the interior subset of array A.

@@

>>>>

%
% \section{Introduction}
%    \subsection{Approach}
%    \subsection{Why Fortran?}
%    \subsection{Comparison to Other Languages}
%

%
% \section{Programming model}
%    \subsection{Fortran Syntax}   
%       \subsubsection*{Array notation}
%       \subsubsection*{Elemental functions}
%       \subsubsection*{Pure procedures}
%       \subsubsection*{Shift functions}
%       \subsubsection*{Regions}
%    \subsection{New Functions}
%    \subsection{Parallelism}
%    \subsection{Limitations}
%

@@

%% NEW - CER
However, the performance that these accelerator architectures offer comes at a
cost as processor architectures are trending toward multiple cores with
instances of integrated accelerator units and less of a reliance on memory
caches.  These changes place a heavy burden on application programmers as they
are forced to adapt to the new systems.  An especially challenging problem
faced by application programmers is not only how to program to these new
architectures (considering the massive scale of concurrency available), but
also how to design programs that are portable across the changing landscape of
computer architectures.  The fundamental question we address is what
programming model and language constructs are best suited for these platforms.

%% NEW - CER
%%%We will examine how the existing data-parallel constructs in Fortran can be combined with coarrays or MPI to provide effectively a new parallel programming language, one that is evolutionary in nature and provides complete compatibility with existing applications and libraries.  Data parallelism is a high-level abstraction that is, at the same time, both easier to program and gives the compiler more leeway (if fully exploited) in retargeting a program to different computer architectures.

@@

We provide (via Fortran interfaces in the ForOpenCL library) a mechanism to
call the C OpenCL runtime and enable Fortran programmers to access OpenCL
kernels.  Transformations are supplied that provide a mechanism for converting
Fortran procedures written in the Fortran subset described in this paper to
OpenCL kernels.  We use the ROSE compiler infrastructure \cite{} to develop
these transformations.  ROSE uses the Open Fortran Parser \cite{} to parse
Fortran 2008 syntax and can output (unparse) to OpenCL.  Since ROSE's
intermediate representation (IR) was constructed to represent multiple
languages, it is relatively straightforward to transform high-level Fortran IR
nodes to OpenCL nodes.  Furthermore, the Fortran array syntax maps directly to
one, two, and three-dimensional thread groups in OpenCL.

Transformations for general Fortran procedures are not provided.  Furthermore,
a mechanism to transform the calling site to automatically invoke
OpenCL kernels is not provided at this time.  While it is possible to
accomplish this task within ROSE, it is considered outside the scope of this
paper.

We examine the performance of the Fortran data-parallel abstraction when
transformed to OpenCL to run on GPU architectures.  Since single node performance
is often given as a reason for not using data-parallel constructs within Fortran,
we consider the performance of serial data-parallel codes compared with the
usage of explicit loop constructs.

We study automatic transformations and the performance for an
application example that is typical of many applications that are based on
finite-difference or finite-volume methods in computational fluid dynamics
(CFD).  The example is a simple shallow water model in two dimensions using
finite volume methods with .... for time update.

%% NEW - CER - copied from proposal
An initial study was made for an important procedure in PAGOSA,
production code at LANL completely written in data-parallel
Fortran.  We investigated automatically transforming this code
to run on LANL's Petaflop Roadrunner computer (a hybrid mixture of Opterons
and IBM Cell processors).  We demonstrated that a source-to-source compiler
can automatically vectorize and parallelize (both within the Cell processor
and across nodes) a small section of this code.  Preliminary results showed a
12 times performance gain on the Cell processor over a traditional single core
processor.  A factor of 9 is gained from vectorizing and parallelizing the
code to run on the Cell and the rest is gained from automatically overlapping
inter-node communication with computation on the Cell.

@@

@@

%% NEW - CER
However, with the introduction of Fortran 90 (and later revisions to the
standard), Fortran is a truly modern programming language. Much of the arcane
baggage associated with it is actually no longer part of the language used by
modern programmers, and exists in a deprecated form simply to support legacy
codes. It is now modular and has many object-oriented features. Furthermore,
Fortran should be of interest to those studying parallel programming because
of its functional and data parallel constructs and because of the coarray
notation introduced after Fortran 2003. Unlike languages like C and C++, which
are considered to be more modern, Fortran has become a truly parallel language
with features added to recent language standards.

Yet, a likely question that one may pose is ``\emph{Why Fortran and not a more
  modern language like X?}''  The recent rise in interest in concurrency and
parallelism at the language level driven by multicore CPUs and manycore
accelerators has driven a number of new language developments, both as novel
languages and extensions on existing ones.

@@

\section{Comparison to Other Languages}

%% Matt you should provide an historical (and current) context here

%% cites: ZPL, CM5 stencil compilers POOMA, HPL OpenMP, CUDA

% put something here on: regions, array notation, ...

\section{Programming Model}

The static analysis and source-to-source transformations used are very
basic but require the programmer to use a subset of Fortran
that employs a data-parallel programming model.  In particular, it
encourages use of language features that were introduced and
standardized in the Fortran 90 language specification.  In this
section we describe the set of Fortran 90 features that our analysis
and transformation method are based on.

@@

%% NEW - CER
Arrays are first class citizens in Fortran as they are part of the language
type system rather than just a pointer to raw memory as in C. A Fortran array
contains metadata associated with it that describes the array's shape and
size. This metadata is lacking in other languages, such as C, C++, and Java.
Even in the case where multidimensional array data structures are available
(such as in C++ or Java), they exist outside the language standard, limiting
the ability of the compiler to analyze them fully. The existence of this
metadata allows the use of array notation whereby explicit indexing of arrays
is not required. For example, if {\tt A}, {\tt B}, and {\tt C} are all arrays
of the same rank and shape and {\tt s} is a scalar, then the statement

 C = A + s*B

results in the sum of elements of {\tt A} and {\tt s} times the elements of
{\tt B} being stored in the corresponding elements of {\tt C}. The first
element of {\tt C} will contain the value of the first element of {\tt A}
added to the first element of {\tt c*B}.  Note that no explicit iteration over
array indices is needed and that the individual operators, plus, times, and
assignment are applied by the compiler to individual elements of the arrays
independently.  Thus the compiler is able to spread the computation in the
example across any hardware threads under its control.

%% NEW - CER
%%% The use of array notation allows one to program in a data-parallel subset of Fortran. This style of programming makes use of array notation and pure and elemental functions to operate on array elements.

%% NEW - CER
%%% Because elemental functions return scalar values and are free from side effects, the compiler is free to distribute the computation over any hardware processing elements available to it, such as the multiple cores and vector units on an Intel or AMD processor or the Synergistic Processing Units on an IBM Cell processor.

%% NEW - CER
An example of shown above using array syntax could be refactored to use
the elemental function,

\begin{verbatim}
   pure elemental real function elem_example(a, b, s)
      real, intent(in) :: a, b, s
      elem_example = a + s*b
   end function
\end{verbatim}

and called with

\begin{verbatim}
   C = elem_example(A, B, s)
\end{verbatim}

Note that while {\tt elem_example} is defined in terms of purely scalar
quantities, it can be \emph{applied} to arrays as shown.

While this may seem like a trivial example, such simple functions may be
composed with other elemental functions to perform powerful computations,
especially when applied to arrays.  ForOpenCL transforms elemental functions to inline OpenCL
functions.  Thus there is no penalty for usage of elemental functions and provide
a convient mechanism to express algorithms in simpler segments.

Also note that the plus and times operators shown in the data-statement
example can be seen as an elemental functions, as plus and times are defined
in terms of scalars but can be applied to whole arrays and can return an array
result.

Pure procedures, like elemental functions, must be free of side effects.
(athough it may change the value of array elements in an array argument,
unlike elemental arguments which must have the intent(in) attribute).
The absence of side effects removes ordering constraints from the compiler
allowing it to invoke pure functions out of order.  Procedures and functions
of this sort are also common in pure functional languages like Haskell,
and are exploited by compilers in order to emit parallel code automatically
due to their suitability for compiler-level analysis.

Since pure procedures don't have side effects they a candidates for running on
accelerators in OpenCL.  Currently ForOpenCL transforms pure procedures to
OpenCL kernels that emph{do not} call other procedures, except for elemental
functions.

\subsubsection*{Shift functions}

\subsubsection*{Regions}

Borrowing from ZPL, we introduce the concept of regions to Fortran.  In
Fortran, regions are expressed as pointers to a subsection of an existing array.
Regions may refer to an interior portion of an array or the entire array.

The use of regions implies the use of the halo (or ghost) cell pattern 
where the size of an array is increased to provide extra array elements
surrounding the interior portion of the array.  
%parlab.eecs.berkeley.edu/wiki/_media/patterns/ghostcell.pdf

Regions are similar to the shift operator as they can be used to reference
portions of the array that are shifted with respect to the interior portion.
However, unlike the shift operator, regions are not expressed in terms of
boundary conditions and thus don't explicitly \emph{require} a knowledge of
not the application of boundary conditions locally.  Thus, as will be shown
below, regions are more suitable for usage by OpenCL thread groups which
access only local subsections of an array stored in global memory.

Unfortunately, as regions functions return pointers to arrays, they cannot
be used in pure procedures in current Fortran.  We relax this constraint in
ForOpenCL for region pointers only.

\subsection{New Functions}

% TODO - Craig - describe these functions
%   region
%   interior (can be written in terms of regions) but also specifies the thread
%       group domain)
%   more ...
%

\subsection{Parallelism}

% TODO - Matt - do you want a try at this.

% need to describe processor hierarchy of nodes and accelerators
% need to describe memory hierarchy as well in regards to global array view,
% node array view (with a copy is GPU global memory) and local GPU memory which
% further subdivides the node array memory

% figures on slicing and dicing of memory region go here

There are several advantages to this style of programming using array syntax, shifts, regions,
and pure and elemental functions:

\begin{itemize}
	\item There are no loops or index variables to keep track of.  Off by one
index errors are a common programming mistake.
        \item The written code is closer to the algorithm, easier to understand, and is usually substantially shorter.
	\item Semantically the intrinsic functions return arrays by value (though not regions).  This is usually what the algorithm requires.
	\item Because pure and elemental function are free from side effects, it is easier for a compiler to schedule the work to be done in parallel.
\end{itemize}

@@

%% NEW - CER
%%% The power in this notation is that the compiler is free to distribute the computation in these expressions over any hardware processing elements available to it, such as the vector units on an Intel or AMD processor or the Synergistic Processing Units on an IBM Cell processor.  If the arrays are declared as coarrays, this includes spreading the computation over \emph{all of the nodes in   a cluster as well.}  In this case, communication occurs within the \texttt{cshift} functions, though the compiler is free to overlap communication and computation by scheduling the communication early and performing the computation on the interior of the arrays while waiting for the communication to complete.

%% NEW - CER
Complete and very concise and elegant programs can be built with procedures
similar to the example shown above. To aid this effort, Fortran
supplies intrinsic functions like the array constructors (\texttt{CSHIFT},
\texttt{EOSHIFT}, \texttt{MERGE}, \texttt{TRANSPOSE}, ...), the array location
functions (\texttt{MAXLOC} and \texttt{MINLOC}), and the array reduction
functions (\texttt{ANY}, \texttt{COUNT}, \texttt{MINVAL}, \texttt{SUM},
\texttt{PRODUCT}, ...).  To this set we add region functions described above.

%% NEW - CER
This style of programming meets the requirements we have set for a programming
model for developing applications suitable for acceleration.  It allows the programmer to
program at a very-high level of abstraction while providing the compiler with
maximum flexibility in targeting the application for a particular hardware
architecture.  The data parallel programming model simultaneously meets the
seemingly conflicting goals of maintainability, portability, and performance.

@@

%% NEW - CER
A goal of this paper is to demonstrate that parallel programs written in
this style of Fortran are maintainable and can achieve good performance on a
variety of processor architectures, including possibly, future terascale
architectures from Intel (such as the XXX architecture).

@@

\section{Source-To-Source Transformations}

% TODO - craig - I need to go through this section and replace PetaVision
% with the shallow water code

@@

This sections show short Fortran code examples and OpenCL equivalent.
The notation uses uppercase for arrays and lowercase for scalar quantities.
Interior array subsections are denoted by an i succeeding the array, e.g. Ai
is the interior region of array A.

@@

\section{Conclusions}

%% Explain what we have accomplished (copied from proposal)

%% NEW - CER
The goal of this work has been to exploit source-to-source transformations that
allow programmers to develop and maintain programs at a high-level of
abstraction, without coding to a specific hardware architecture.
Furthermore these transformations allow multiple hardware architectures
to be targeted without changing the high-level source.  It also removes the
necessity for application programmers to understand details of the accelerator
architecture or to know OpenCL.

% Summarize results
